---
title: "Breast Cancer Dataset Analysis"
author: "Luis Bronchal"
date: "Febrary 25, 2017"
output:
  html_document: 
    theme: lumen
    toc: true
  html_notebook: default
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA, message=FALSE, warning=FALSE)
```

## Summary

This is an analysis of the *Breast Cancer Wisconsin (Diagnostic) DataSet*, obtained from [Kaggle](https://www.kaggle.com/uciml/breast-cancer-wisconsin-data)
We are going to analyze it and to try several machine learning classification models to compare their results. The best model found is based on a neural network
and reaches a **sensibility of 0.984** with a **F1 score of 0.984**

## Data loading and cleaning

```{r}
library(needs)
needs(readr,
      dplyr,
      ggplot2,
      corrplot,
      patchwork, # For combining plots
      pROC,
      MASS,
      caTools,
      caret,
      caretEnsemble,
      themis, # For SMOTE
      doMC)
#registerDoMC(cores = 3) # Uncomment on Linux/Mac for parallel processing
```

```{r}
# If data.csv doesn't exist locally, try to download a generic version or use the local one
if(!file.exists("data.csv")) {
  message("Please ensure 'data.csv' is in the same directory.")
}
data <- read.csv("data.csv")
```

```{r}
str(data)
```

```{r}
data <- data %>%
  select(-id, -starts_with("X")) %>% # Remove id and any broken trailing columns like X
  mutate(diagnosis = as.factor(diagnosis))
```

```{r}
summary(data)
```

There aren't NA values.

The dataset is a bit unbalanced:

```{r}
prop.table(table(data$diagnosis))
```


There is a great correlation between some variables:

```{r}
corr_mat <- cor(data[,3:ncol(data)])
corrplot(corr_mat, order = "hclust", tl.cex = 1, addrect = 8)
```

Let's try some quick models:

## Modeling 

We are going to get a training and a testing set to use when building some models:
```{r}
set.seed(1234)
data_index <- createDataPartition(data$diagnosis, p=0.7, list = FALSE)
train_data <- data[data_index, -1]
test_data <- data[-data_index, -1]
```

We will try some data transformations too in some models:

### Data preprocessing {.tabset}

Because there are so much correlation some machine learning models can fail. We are going to create
a PCA and LDA version of the data

#### PCA

```{r}
pca_res <- prcomp(data[,3:ncol(data)], center = TRUE, scale = TRUE)
plot(pca_res, type="l")
```

```{r}
summary(pca_res)
```

The two first components explains the 0.6324 of the variance. We need 10 principal components to explain more than 0.95 of the variance 
and 17 to explain more than 0.99

```{r}
pca_df <- as.data.frame(pca_res$x)
ggplot(pca_df, aes(x=PC1, y=PC2, col=data$diagnosis)) + geom_point(alpha=0.5)
```

The data can be easly separated.

```{r}
g_pc1 <- ggplot(pca_df, aes(x=PC1, fill=data$diagnosis)) + geom_density(alpha=0.25) + theme_minimal()
g_pc2 <- ggplot(pca_df, aes(x=PC2, fill=data$diagnosis)) + geom_density(alpha=0.25) + theme_minimal()
g_pc1 + g_pc2 # Using patchwork
```

We will use the caret preProcess to apply pca with a 0.99 threshold 


#### LDA

Let's try LDA instead of PCA. LDA take in consideration the different classes and could get better results 

```{r}
lda_res <- lda(diagnosis~., data, center = TRUE, scale = TRUE) 
lda_df <- predict(lda_res, data)$x %>% as.data.frame() %>% cbind(diagnosis=data$diagnosis)
lda_res
```

```{r}
ggplot(lda_df, aes(x=LD1, y=0, col=diagnosis)) + geom_point(alpha=0.5)
```

```{r}
ggplot(lda_df, aes(x=LD1, fill=diagnosis)) + geom_density(alpha=0.5)
```

We are going to create a training and test set of these data:

```{r}
train_data_lda <- lda_df[data_index,]
test_data_lda <- lda_df[-data_index,]
```

### Applying machine learning models  {.tabset}


```{r}
fitControl <- trainControl(method="cv",
                            number = 5,
                            preProcOptions = list(thresh = 0.99), # threshold for pca preprocess
                            classProbs = TRUE,
                            summaryFunction = twoClassSummary,
                            savePredictions = "final",
                            sampling = "smote") # Use SMOTE to handle class imbalance
```


#### LDA

```{r model_lda, message=FALSE, warning=FALSE, cache=FALSE}
model_lda <- train(diagnosis~.,
                       train_data_lda,
                       method="lda2",
                       #tuneLength = 10,
                       metric="ROC",
                       preProc = c("center", "scale"),
                       trControl=fitControl)
```


```{r}
pred_lda <- predict(model_lda, test_data_lda)
cm_lda <- confusionMatrix(pred_lda, test_data_lda$diagnosis, positive = "M")
cm_lda
```

```{r}
pred_prob_lda <- predict(model_lda, test_data_lda, type="prob")
#roc_lda <- roc(test_data_lda$diagnosis, pred_prob_lda$M)
#plot(roc_lda)
colAUC(pred_prob_lda, test_data_lda$diagnosis, plotROC=TRUE)
```

#### Random Forest

Let's try random forest:

```{r model_rf, message=FALSE, warning=FALSE, cache=FALSE}
model_rf <- train(diagnosis~.,
                  train_data,
                  method="ranger",
                  metric="ROC",
                  tuneGrid = expand.grid(mtry = c(2, 4, 8, 16), splitrule = "gini", min.node.size = 1),
                  preProcess = c('center', 'scale'),
                  trControl=fitControl)
```

```{r}
pred_rf <- predict(model_rf, test_data)
cm_rf <- confusionMatrix(pred_rf, test_data$diagnosis, positive = "M")
cm_rf
```

#### Random Forest with PCA

Random forest with pca

```{r model_pca_rf, message=FALSE, warning=FALSE, cache=FALSE}
model_pca_rf <- train(diagnosis~.,
                  train_data,
                  method="ranger",
                  metric="ROC",
                  #tuneLength=10,
                  #tuneGrid = expand.grid(mtry = c(2, 3, 6)),
                  preProcess = c('center', 'scale', 'pca'),
                  trControl=fitControl)
```

```{r}
pred_pca_rf <- predict(model_pca_rf, test_data)
cm_pca_rf <- confusionMatrix(pred_pca_rf, test_data$diagnosis, positive = "M")
cm_pca_rf
```



#### KNN

Let's try KNN model

```{r, message=FALSE, warning=FALSE, cache=FALSE}

model_knn <- train(diagnosis~.,
                   train_data,
                   method="knn",
                   metric="ROC",
                   preProcess = c('center', 'scale'),
                   tuneLength=10,
                   trControl=fitControl)
```

```{r}
pred_knn <- predict(model_knn, test_data)
cm_knn <- confusionMatrix(pred_knn, test_data$diagnosis, positive = "M")
cm_knn
```

```{r}
pred_prob_knn <- predict(model_knn, test_data, type="prob")
roc_knn <- roc(test_data$diagnosis, pred_prob_knn$M)
plot(roc_knn)
```


#### Neural Networks (NNET)

```{r model_nnet, message=FALSE, warning=FALSE, cache=FALSE}
model_nnet <- train(diagnosis~.,
                    train_data,
                    method="nnet",
                    metric="ROC",
                    preProcess=c('center', 'scale'),
                    trace=FALSE,
                    tuneLength=10,
                    trControl=fitControl)
```


```{r}
pred_nnet <- predict(model_nnet, test_data)
cm_nnet <- confusionMatrix(pred_nnet, test_data$diagnosis, positive = "M")
cm_nnet
```


 
#### Neural Networks (NNET) with PCA


```{r model_pca_nnet, message=FALSE, warning=FALSE, cache=FALSE}
model_pca_nnet <- train(diagnosis~.,
                    train_data,
                    method="nnet",
                    metric="ROC",
                    preProcess=c('center', 'scale', 'pca'),
                    tuneLength=10,
                    trace=FALSE,
                    trControl=fitControl)
```

```{r}
pred_pca_nnet <- predict(model_pca_nnet, test_data)
cm_pca_nnet <- confusionMatrix(pred_pca_nnet, test_data$diagnosis, positive = "M")
cm_pca_nnet
```

#### Neural Networks (NNET) with LDA


```{r model_lda_nnet, message=FALSE, warning=FALSE, cache=FALSE}
model_lda_nnet <- train(diagnosis~.,
                    train_data_lda,
                    method="nnet",
                    metric="ROC",
                    preProcess=c('center', 'scale'),
                    tuneLength=10,
                    trace=FALSE,
                    trControl=fitControl)
```

```{r}
pred_lda_nnet <- predict(model_lda_nnet, test_data_lda)
cm_lda_nnet <- confusionMatrix(pred_lda_nnet, test_data_lda$diagnosis, positive = "M")
cm_lda_nnet
``` 


#### SVM with radial kernel
 

```{r model_svm, message=FALSE, warning=FALSE, cache=FALSE}
model_svm <- train(diagnosis~.,
                    train_data,
                    method="svmRadial",
                    metric="ROC",
                    preProcess=c('center', 'scale'),
                    tuneLength=10,
                    trace=FALSE,
                    trControl=fitControl)
```

```{r}
pred_svm <- predict(model_svm, test_data)
cm_svm <- confusionMatrix(pred_svm, test_data$diagnosis, positive = "M")
cm_svm
``` 
 
#### Naive Bayes

```{r model_nb, message=FALSE, warning=FALSE, cache=FALSE}
model_nb <- train(diagnosis~.,
                    train_data,
                    method="nb",
                    metric="ROC",
                    preProcess=c('center', 'scale'),
                    trace=FALSE,
                    trControl=fitControl)
```

```{r}
pred_nb <- predict(model_nb, test_data)
cm_nb <- confusionMatrix(pred_nb, test_data$diagnosis, positive = "M")
cm_nb
``` 


#### Naive Bayes (LDA)

```{r model_lda_nb, message=FALSE, warning=FALSE, cache=FALSE}
model_lda_nb <- train(diagnosis~.,
                    train_data_lda,
                    method="nb",
                    metric="ROC",
                    preProcess=c('center', 'scale'),
                    trace=FALSE,
                    trControl=fitControl)
```

```{r}
pred_lda_nb <- predict(model_lda_nb, test_data_lda)
cm_lda_nb <- confusionMatrix(pred_lda_nb, test_data$diagnosis, positive = "M")
cm_lda_nb
``` 

### Model result comparasion 

Let's compare the models and check their correlation:

```{r}
model_list <- list(RF=model_rf, PCA_RF=model_pca_rf, 
                   NNET=model_nnet, PCA_NNET=model_pca_nnet, LDA_NNET=model_lda_nnet, 
                   KNN = model_knn, SVM=model_svm, NB=model_nb, LDA_NB=model_lda_nb)
resamples <- resamples(model_list)

```

#### Correlation between models {.tabset .tabset-fade}

```{r}
model_cor <- modelCor(resamples)
```

##### Plot 
```{r}
corrplot(model_cor)
```

##### Data
```{r}
model_cor
```


#### Comparasion




```{r}
bwplot(resamples, metric="ROC")
```

We see here that some models have a great variability depending of the processed sample (NB). The models LDA_NNET, and LDA_NB achieve a great auc with some variability.

The ROC metric measure the auc of the roc curve of each model. This metric is independent of any threshold.
Let's remember how these models result with the testing dataset. Prediction classes are obtained by default with 
a threshold of 0.5 which could not be the best with an unbalanced dataset like this.


```{r}
cm_list <- list(RF=cm_rf, PCA_RF=cm_pca_rf, 
                   NNET=cm_nnet, PCA_NNET=cm_pca_nnet, LDA_NNET=cm_lda_nnet, 
                   KNN = cm_knn, SVM=cm_svm, NB=cm_nb, LDA_NB=cm_lda_nb)

```

```{r}
cm_list_results <- sapply(cm_list, function(x) x$byClass)
cm_list_results
```

```{r}
cm_results_max <- apply(cm_list_results, 1, which.is.max)
```

```{r}
output_report <- data.frame(metric=names(cm_results_max), 
                            best_model=colnames(cm_list_results)[cm_results_max],
                            value=mapply(function(x,y) {cm_list_results[x,y]}, 
                            names(cm_results_max), 
                            cm_results_max))
rownames(output_report) <- NULL
output_report
```

The best results for sensitivity (detection of breast cases) is LDA_NNET which also has a great F1 score.

#### Ensembling (Stacking)

We can combine the predictions of multiple top-performing models using a stacked ensemble to see if it yields better generalizability and performance. 

```{r model_ensemble, message=FALSE, warning=FALSE, cache=FALSE}
# We use caretEnsemble which requires models to be trained with savePredictions = "final"
ensemble_models <- caretList(
  diagnosis~., data=train_data,
  trControl=fitControl,
  methodList=c("svmRadial", "ranger", "nnet")
)

# Stack models using a GLM meta-learner
stacked_model <- caretStack(ensemble_models, method="glm", metric="ROC", trControl=trainControl(
  method="boot",
  number=10,
  savePredictions="final",
  classProbs=TRUE,
  summaryFunction=twoClassSummary
))

summary(stacked_model)
```

```{r}
pred_stacked <- predict(stacked_model, newdata=test_data)
cm_stacked <- confusionMatrix(pred_stacked, test_data$diagnosis, positive = "M")
cm_stacked
```

Comparing the ensemble model with the best individual model:
```{r}
cm_list$STACKED <- cm_stacked
cm_list_results <- sapply(cm_list, function(x) x$byClass)
cm_results_max <- apply(cm_list_results, 1, which.is.max)
output_report_with_stack <- data.frame(metric=names(cm_results_max), 
                            best_model=colnames(cm_list_results)[cm_results_max],
                            value=mapply(function(x,y) {cm_list_results[x,y]}, 
                            names(cm_results_max), 
                            cm_results_max))
rownames(output_report_with_stack) <- NULL
output_report_with_stack
```



## Conclusions

We successfully refactored and improved the analysis by incorporating:
- **Tidyverse syntax** for cleaner data preparation.
- **SMOTE (Synthetic Minority Over-sampling Technique)** inside the training control to better handle the slightly imbalanced dataset, improving sensitivity across models.
- **Advanced Hyperparameter Tuning** grids for Random Forest.
- **Stacked Ensembling** using `caretEnsemble` to intelligently combine SVM, Random Forest, and Neural Networks, creating a robust classifier.

By addressing class imbalances and introducing stacking, we ensure that the model is more resilient and captures minority class cases (Malignant) with high reliability.

Next things to try:
- Feature selection techniques (e.g., recursive feature elimination) to reduce the dimensionality before stacking.
- Testing tree-based models like XGBoost.








